Manuscript

### 2025-03-05 Manuscript_MD_ScoreReLiability.qmd
update manuscript intro

---
title: "Manuscript"
format: docx
editor: 
  markdown: 
    wrap: 72
---

@@

# Score ReLiability – Can we explain effect size heterogeneity using score reliability?

Large-scale collaborative replication efforts have sparked discussions
surrounding the replicability of psychological phenomena. However, for
attempts to estimate a phenomenon’s replicability or to predict whether
a future replication will replicate successfully, heterogeneity is a
crucial parameter that needs to be assessed. In the meta-analytic
context, heterogeneity describes the variation in effect sizes, free of
sampling error, if these are not stable or constant across replications.
Therefore, the presence of heterogeneity implies that some replications
of a single phenomenon may be successful, while others are not. If
heterogeneity grows, meaning the phenomenon’s effect size varies more
strongly for unexplained reasons, the probability of observing an effect
size around zero or even in the negative space grows larger as well. If
we know the mean size and heterogeneity of a phenomenon’s effect size,
we would theoretically be able to establish a baseline of expected
replication rate (reference).

Similarly, it has been argued that heterogeneity in effect sizes is an
indicator of the theory’s “completeness” surrounding the phenomenon.
Linden and Hönekopp argue that “low (as opposed to high) heterogeneity
reflects a more advanced understanding of the subject matter being
studied” (2021, p.2). Similarly, Schuetze and von Hippel (2023) argue
that heterogeneity in effects is an indicator of a vague, poorly
specified theory.

Re-analyses of multi-site direct replications have demonstrated that, if
a psychological effect size can be distinguished from zero,
heterogeneity is present across most phenomena (ref). Large-scale
attempts of direct replications, using identical protocols, such as the
Many Labs studies or Registered Replication Reports (references), for
the first time allow researchers to estimate heterogeneity undistorted
by differences in experimental designs. In re-analyses of these studies,
Olsson-Collentine et al. (2020) identify a positive correlation between
a phenomenon’s effect size and its heterogeneity. Similarly, van Erp et
al. (2017) or Stanley et al. (2018) estimate strong degrees of
heterogeneity across conceptual replications in Psychology. In a
separate re-analysis of large scale direct replications, Renkewitz et
al. (2025) identify substantial heterogeneity in almost all projects
where a non-zero effect could be identified. This aligns with the
correlation found by Olsson-Collentine et al. (2020).

@@

The claims made by W&D imply that corrections of individual observed ESs
for their unreliability should lead to lower estimates of heterogeneity
in meta-analyses of corrected ES. If heterogeneity in score reliability
adds to ES heterogeneity, attenuation correction procedures should
eliminate that additional variation, as all corrected ES come with
identical (perfect) score reliability.

Re-analyses of the collaborative multi-site replication efforts have
uncovered differences in score reliability across replications of the
same phenomena (ref McShaw etc.). The claims made in W&D and H&S imply
that the ES heterogeneity uncovered in these projects therefore could
potentially be explained by these differences in score reliability. The
same argument has been proposed by both original authors of phenomena
that did not replicate or carry substantial heterogeneity (references)
as well as the authors of the multi-site replication attempts (refs).
Over the following pages, we assess for the first time, whether the
differences in score reliability do indeed explain, and thereby reduce,
the ES heterogeneity observed across the different multi-site
replication efforts. Even though only a limited number of phenomena
allow for such an analysis, we find no evidence for a reduction in ES
heterogeneity after taking score reliability into account. Subsequently,
we discuss why the expectation formulated in H&S does not allow for such
claims. Alternatively, we explore to what extent defining ES as a random
ratio variable helps understand how ES heterogeneity is affected by
differences in score reliability. We close with a discussion on what
future meta-analysts can expect, when dealing with issues of
(un)reliability in assessments of meta-analyses of ES.

@@

As all phenomena in the collection fulfil the first condition, 50
phenomena have been catalogued where the effect is studied by comparing
the outcome across two separate groups. While the majority of designs
make use of some form of control-treatment manipulation, randomly
assigning participants to a condition (20 phenomena), some phenomena
discuss the effect of pre-existing differences, e.g. biological sex (2
phenomena). From the collection of 50 phenomena, 22 fulfil the second
condition, so that estimates of score reliability can be derived. For
these phenomena, the dependent variable used to construct the effect
size is measured by employing more than a single indicator. At the same
time, however, this implies that the remaining 28 phenomena did not make
use of more than a single indicator to measure the dependent variable.
Therefore, for the majority of phenomena it is not possible to assess in
how far measurement quality is sufficiently high in terms of e.g.
internal consistency or low random error variance.

From the 22 phenomena that employed more than a single indicator, 19
made use of Likert-style items, where respondents would indicate their
agreement to some kind of statement. The scale-length varied from 3 to
10-points. The items measuring the remaining 3 phenomena were coded
dichotomously, as responses were right or wrong (2 cases), or expressed
agreement vs disagreement dichotomously (1 case). Subsequently, those
phenomena need to be identified where the meta-analytic mean effect size
is statistically significant different from zero. This will be done as a
first step of this manuscript’s analysis procedure. Detailed information
concerning the different phenomena and how they were measured can be
found at \[osf-link\], while a brief summary can be found in Table A.

@@

@eq-HS implies that heterogeneity in $\delta_0$ is essentially a
function where heterogeneity and mean value of $\delta$ are weighted by
mean score reliability $\rho_{xx‘}$ and its heterogeneity. While it may
not be self-evident in the equation itself, it is easy to construct
cases where heterogeneity in $\delta_0$ is inflated or deflated by
heterogeneity in $\rho_{xx‘}$, contradicting Hunter & Schmidt’s
interpretation of this equation. In @fig-HS, the results of employing
@eq-HS to compute the heterogeneity (variance) in corrected ES is
highlighted. All else held equal, only the individual level of mean
reliability $\mu_{\rho_{xx‘}}$ is varied from .5 to .8. Mean corrected
ES $\mu_{\delta}$ is held constant at 1, its variance $\tau^2_{\delta}$
held constant at .04 and the variance of score reliability
$\tau^2_{\rho_{xx‘}}$ held constant at .02.

```{r, echo = FALSE}
#| label: fig-HS
#| fig-cap: Implications by Hunter & Schmidt's @eq-HS

@@

In @fig-HS, the dashed red line corresponds to the variance of corrected
ES.

If everything else is held constant, the variance in uncorrected ES can
be a function only of the mean level of score reliability
$\mu_{\rho_{xx‘}}$, according to H&S equation. In that case, Figure 2
demonstrates that correcting for imperfect score reliability does not
inevitably lead to lower heterogeneity, as the line representing
heterogeneity in uncorrected ES does cross the red line, representing
the heterogeneity in corrected ES. This implies that, as long as mean
score reliability is below .72, an attenuation correction would lead to
a reduction in ES heterogeneity. As mean score reliability grows larger
than .72, however, the heterogeneity in uncorrected ES grows larger than
the heterogeneity in corrected ES - an attenuation correction would
therefore only increase the extent of ES heterogeneity. This
demonstrates that, depending on the mean level of score reliability,
heterogeneity in uncorrected ES can be higher or lower than
heterogeneity in corrected ES. As heterogeneity in corrected ES is not
always smaller, the interpretationsof H&S concerning this equation is
erroneous.

However, @eq-HS is additionally flawed, as it violates a crucial
definition. As Hunter & Schmidt point out, @eq-HS is only valid “If the
level of reliability is independent of the true effect size across
studies \[…\]” (2014, p. 309). However, both parameters, ES $\delta$ and
score reliability ρxx‘, can not be independent parameters by definition.
ES $\delta$ is the “true” standardized mean difference, unaffected by
sampling or measurement error. The ES $\delta_0$ is standardized using
the total standard deviation, which is inflated by measurement error.
Equation 1, demonstrating how Cohen’s d is computed demonstrates how the
mean difference is standardized in this case. As demonstrated further
above in @eg-dct, ES $\delta$ is standardized using the true standard
deviation in the sense of CTT. The fact that the true standard deviation
parameter $\sigma_T$ is found both in @eq-dct and @eq-rel demonstrates
that ES $\delta$ and score reliability $\rho_{xx‘}$ can not be
independent variables. Therefore, the basic premise of @eq-HS is not
fulfilled, as the assumption of variable independence could not be
fulfilled.

@@

Instead, what is guaranteed is that the expected value of $\sigma_T$ is
smaller than the expected value of $\sigma_X$. Similarly, the
heterogeneity in $\sigma_T$ is guaranteed to be smaller than the
uncorrected heterogeneity in $\sigma_X$. However, since the expected
value of either standard deviation is placed in the denominator of the
function and the variance of either is placed in the numerator of the
function, @eq-HS is not sufficient to explain how heterogeneity in ES
changes due to the attenuation correction. Instead, we introduce two
additional metrics.

@@

The data-sets available to this project fall into two categories: direct
multi-site replications of a specific phenomenon, employing identical
protocols across samples; a large-scale collaborative effort to assess a
single phenomenon globally, distributed across a large number of
independent laboratories. For both categories, rather arbitrary sampling
schemes were used. The majority of respondents are students at the
respective universities, taking part in the studies voluntarily or
because they were required as part of their study's programme (at times
compensation was awarded). This is not the same as random sampling.
These samples are put together arbitrarily, it is unlikely that the
underlying distribution of true scores is meaningful for a properly
specified population. The between-studies randomness introduced by these
undefined sampling schemes manifests itself in true score variance
heterogeneity $\tau^2_{\sigma^2_X}$. Between-studies randomness
introduces by issues of measurement quality manifest themselves in error
score variance heterogeneity $\tau^2_{\sigma^2_X }$. This implies that
it should be of little surprise that the differences in true score
variance largely exceed those in error score variance.

Without sampling schemes that assure that the samples are constructed to
be representative for some kind of population, we have no reason to
assume that there should be few differences in the dispersion of true
scores. Even though the measuring instruments are translated to
different languages, which is bound to introduce differences in how much
random noise is added to scores across samples, we have no reason to
assume that these differences are larger than those introduced by
employing no sampling schemes. Therein, we believe that, as the
differences in score reliability are largely introduced by failure of
ensuring appropriate sampling mechanisms, not by differences in how well
the instrument captures the underlying construct, attenuation correction
can not be expected to remedy ES heterogeneity.

However, this also means that results contrary to what we observed in
this project are not unthinkable. Valid sampling schemes need to be
considered, ensuring that the samples are representative for populations
that can be meaningfully compared. In that case, we are convinced that
true score variance should be more stable across samples and differences
in the instrument's measuring quality are bound to have a much larger
impact. Not only would an attenuation correction in such a case actually
have the potential of reducing ES heterogeneity, the variability
underlying a psychological phenomenon could be meaningfully assessed
across samples - something that would greatly help attempts to identify
universal and non-universal patterns of psychological phenomena. As it
stands, current approaches to multi-site direct replications and other
collaborative efforts suffer strongly from the lack of appropriate
sampling schemes.

@@

The analytical arguments presented following Equation 17 rested on the
use of the delta-method, as Equation 17 contained parameters of the
score standard deviations’ distributions, while the $R_1$ and $R_2$
metrics were designed to discuss score variances. The delta-method rests
on a number of assumptions: (1) approximations derived from the
delta-method require for the transformed variable to follow a marginal
normal distribution (ref). Since the variable under discussion here is
the standard deviation, which is bounded to be larger than zero and has
a non-normal sampling distribution (ref), this assumption is most likely
violated; (2) the derivative of the transformation function needs to be
available. As the transformation function is essentially just the square
root, this assumption should hold up; (3) as the delta-method
approximates the parameters of the transformed variable using the
first-order Taylor-expansion, this implies that higher-order terms are
required to be negligible. In practice, this implies that the variance
of the untransformed variable needs to be small. However, as this entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability, this
assumption is probably violated as well. The violation of these
assumptions likely introduce a bias in estimates derived from these
approximations (ref). However, the inequality derived with use of the
delta-method is not used to actually derive any estimates, but to
discuss how an attenuation correction procedure can affect ES
heterogeneity. Therefore, we believe that the violation of these
assumptions is defensible in this case, as the analytical arguments
derived from the inequality should still hold up, nevertheless.
Additionally, we used the empirical examples available to assess in how
far estimates derived using the delta-method approximations diverge from
estimates derived without it. The differences identified are rather
miniscule at the third decimal, reinforcing our belief that these
assumption violations are defensible.

>>>>

---
title: "Manuscript"
format: html
editor: 
  markdown: 
    wrap: 72
---

@@

# Liability or reliability? Exploring the role of effect size attenuation on meta-analytic heterogeneity

Large-scale collaborative replication efforts have sparked discussions
surrounding the replicability of psychological phenomena. For attempts
to estimate a phenomenon’s replicability or to predict whether a future
replication will replicate successfully, heterogeneity is a crucial
parameter that needs to be assessed. In the meta-analytic context,
heterogeneity describes the variation in effect sizes, free of sampling
error, if these are not constant across replications. Therefore, the
presence of heterogeneity implies that some replications of a single
phenomenon may be successful, while others are not. If heterogeneity
grows, meaning the phenomenon’s effect size varies more strongly for
unexplained reasons, the probability of observing an effect size around
zero or even in the negative space grows larger as well. If we know the
mean size and heterogeneity of a phenomenon’s effect size, we would
theoretically be able to establish an expected replication rate
(reference).

Similarly, it has been argued that heterogeneity in effect sizes is an
indicator of the theory’s “completeness” surrounding the phenomenon.
Linden and Hönekopp argue that “low (as opposed to high) heterogeneity
reflects a more advanced understanding of the subject matter being
studied” (2021, p.2). Alternatively, Schuetze and von Hippel (2023)
argue that heterogeneity in effects is an indicator of a vague, poorly
specified theory.

Large-scale attempts of direct replications, using identical protocols,
such as the Many Labs studies or Registered Replication Reports
(references), for the first time allow researchers to estimate
heterogeneity undistorted by selection processes induced by publication
bias. In re-analyses of these studies, Olsson-Collentine et al. (2020)
identify a positive correlation between a phenomenon’s effect size and
its heterogeneity. Similarly, van Erp et al. (2017) or Stanley et al.
(2018) estimate strong degrees of heterogeneity across conceptual
replications in Psychology. In a separate re-analysis of large scale
direct replications, Renkewitz et al. (2025) identify substantial
heterogeneity in almost all projects where a non-zero effect could be
identified. This aligns with the correlation found by Olsson-Collentine
et al. (2020).

One source for this heterogeneity, which is not a typical moderator tied
to a specific theory, might be impairments in score reliability which
vary in their extent across replications. This is an argument put
forward repeatedly by original researchers and replicators alike.
Literature discussing the role of score reliability for standardized
effect sizes claims that differences in score reliability should
inflate, and therefore increase, heterogeneity. Over the following
pages, we assess whether score reliability can in fact be made
responsible for heterogeneity identified across phenomena. Even though
only a limited number of phenomena allow for such an analysis, we find
no empirical evidence for the argument that differences in score
reliability inflate effect size heterogeneity. Therefore, in the last
part of this text we turn towards an alternative discussion using
analytical arguments on how differences in measurement quality affect
standardized effect sizes and their differences.

@@

The claims made in W&D imply that corrections of individual observed ESs
for their unreliability should lead to lower estimates of heterogeneity
in meta-analyses of corrected ES. If heterogeneity in score reliability
adds to ES heterogeneity, attenuation correction procedures should
eliminate that additional variation, as all corrected ES come with
identical (perfect) score reliability.

Re-analyses of the collaborative multi-site replication efforts have
uncovered differences in score reliability across replications of the
same phenomena (ref McShaw etc.). The claims made in W&D and H&S imply
that the ES heterogeneity uncovered in these projects therefore could
potentially be explained by these differences in score reliability.
Similarly, attempting to explain the substantial extent of ES
heterogeneity identified in psychological phenomena, both the original
authors as well as the multi-site replication authors claim that
differences in score reliability might be responsible (references).

Therefore, in the following section we assess to what extent this
intuition can be . The re-analysis of archival data is

@@

As all phenomena in the collection fulfil the first condition, 50
phenomena have been catalogued where the effect is studied by comparing
the outcome across two separate groups. From the collection of 50
phenomena, 22 fulfil the second condition, so that estimates of score
reliability can be derived. For the 22 phenomena, the dependent variable
used to construct the effect size is measured by employing more than a
single indicator. At the same time, however, this implies that the
remaining 28 phenomena did not make use of more than a single indicator
to measure the dependent variable. Therefore, for the majority of
phenomena it is not possible to assess in how far measurement quality is
sufficiently high in terms of e.g. internal consistency or low random
error variance.

From the 22 phenomena that employed more than a single indicator, 19
made use of Likert-style items, where respondents would indicate their
agreement to some kind of statement. The scale-length varied from 3 to
10-points. The items measuring the remaining 3 phenomena were coded
dichotomously, as responses were right or wrong (2 cases), or expressed
agreement vs disagreement dichotomously (1 case). While the majority of
designs make use of some form of control-treatment manipulation,
randomly assigning participants to a condition (20 phenomena), some
phenomena discuss the effect of pre-existing differences, e.g.
biological sex (2 phenomena). Subsequently, those phenomena need to be
identified where the meta-analytic mean effect size is statistically
significant different from zero. This will be done as a first step of
this manuscript’s analysis procedure. Detailed information concerning
the different phenomena and how they were measured can be found at
\[osf-link\], while a brief summary can be found in Table 1.

@@

@eq-HS implies that heterogeneity in $\delta_0$ is essentially a
function where heterogeneity and mean value of $\delta$ are weighted by
mean score reliability $\rho_{xx‘}$ and its heterogeneity. While it may
not be self-evident in the equation itself, it is easy to construct
cases where heterogeneity in $\delta_0$ is inflated or deflated by
heterogeneity in $\rho_{xx‘}$, contradicting Hunter & Schmidt’s
interpretation of this equation. In @fig-hs, the results of employing
@eq-HS to compute the heterogeneity (variance) in corrected ES is
highlighted. All else held equal, only the individual level of mean
reliability $\mu_{\rho_{xx‘}}$ is varied from .5 to .8. Mean corrected
ES $\mu_{\delta}$ is held constant at 1, its variance $\tau^2_{\delta}$
held constant at .04 and the variance of score reliability
$\tau^2_{\rho_{xx‘}}$ held constant at .02.

```{r, echo = FALSE}
#| label: fig-HS
#| fig-cap: Implications by Hunter & Schmidt's @eg-HS

@@

In @fig-hs, the red line corresponds to the variance of corrected ES.

If everything else is held constant, the variance in uncorrected ES can
be a function only of the mean level of score reliability
$\mu_{\rho_{xx‘}}$, according to H&S equation. In that case, Figure 2
demonstrates that correcting for imperfect score reliability does not
inevitably lead to lower heterogeneity, as the line representing
heterogeneity in uncorrected ES does cross the red line, representing
the heterogeneity in corrected ES. Depending on the mean level of score
reliability, heterogeneity in uncorrected ES can be higher or lower than
heterogeneity in corrected ES.

However, @eq-hs is additionally flawed, as it violates a crucial
definition. As Hunter & Schmidt point out, @eq-hs is only valid “If the
level of reliability is independent of the true effect size across
studies \[…\]” (2014, p. 309). However, both parameters, ES $\delta$ and
score reliability ρxx‘, can not be independent parameters by definition.
ES $\delta$ is the “true” standardized mean difference, unaffected by
sampling or measurement error. The ES $\delta_0$ is standardized using
the total standard deviation, which is inflated by measurement error.
Equation 1, demonstrating how Cohen’s d is computed demonstrates how the
mean difference is standardized in this case. As demonstrated further
above in @eg-dct, ES $\delta$ is standardized using the true standard
deviation in the sense of CTT. The fact that the true standard deviation
parameter $\sigma_T$ is found both in @eq-dct and @eq-rel demonstrates
that ES $\delta$ and score reliability $\rho_{xx‘}$ can not be
independent variables. Therefore, the basic premise of @eq-hs is not
fulfilled, as the assumption of variable independence could not be
fulfilled.

@@

Instead, what is guaranteed is that the expected value of $\sigma_T$ is
smaller than the expected value of $\sigma_X$. Similarly, the
heterogeneity in $\sigma_T$ is guaranteed to be smaller than the
uncorrected heterogeneity in $\sigma_X$. However, since the expected
value of either standard deviation is placed in the denominator of the
function and the variance of either is placed in the numerator of the
function, @eq-hs is not sufficient to explain how heterogeneity in ES
changes due to the attenuation correction. Instead, we introduce two
additional metrics.

@@

@@

Paragraph on implications -\> data collected using no specific sampling
schemes - little surprise that the radomness induced by sampling
schemes, which manifests itself in true score variance heterogeneity,
largely surpasses the heterogeneity induced by imperfect measurement.

The analytical arguments presented following Equation 17 rested on the
use of the delta-method, as Equation 17 contained parameters of the
score standard deviations’ distributions, while the $R_1$ and $R_2$
metrics were designed to discuss score variances. The delta-method rests
on a number of assumptions: (1) approximations derived from the
delta-method require for the transformed variable to follow a marginal
normal distribution (ref). Since the variable under discussion here is
the standard deviation, which is bounded to be larger than zero and has
a non-normal sampling distribution (ref), this assumption is most likely
violated; (2) the derivative of the transformation function needs to be
available. As the transformation function is essentially just the square
root, this assumption should hold up; (3) as the delta-method
approximates the parameters of the transformed variable using the
first-order Taylor-expansion, this implies that higher-order terms are
required to be negligible. In practice, this implies that the variance
of the untransformed variable needs to be small. However, as this entire
procedure discusses the influence of variance in standard deviations on
ES heterogeneity, introduced by differences in score reliability, this
assumption is probably violated as well. The violation of these
assumptions likely introduce a bias in estimates derived from these
approximations (ref). However, the inequality derived with use of the
delta-method is not used to actually derive any estimates, but to
discuss how an attenuation correction procedure can affect ES
heterogeneity. Therefore, we believe that the violation of these
assumptions is defensible in this case, as the analytical arguments
derived from the inequality should still hold up, nevertheless.
